---
output# themefurmanu

The `themefurmanu` package provides a comprehensive ggplot2 theme and color palette system designed for Furman University's visual identity. Unlike traditional color palette packages that require you to remember different function names, **themefurmanu uses a unified approach** where you specify the palette type and behavior through arguments to a single set of functions.

> **Note**: This is not an official package from Furman University. It is a personal project to help with visual consistency in data visualizations across teams at Furman that use R.

## Key Features

- **Unified color scale functions**: Use `scale_color_furmanu()` and `scale_fill_furmanu()` for all color needs
- **Intelligent palette system**: Specify discrete vs. continuous, palette type, and special options through arguments
- **Multiple theme variants**: Standard, presentation, and minimal themes
- **Accessibility tools**: Built-in contrast checking and accessible color combinations
- **Export capabilities**: Export palettes to CSS, JSON, RGB, and other formats
- **Automatic font handling**: IBM Plex fonts with graceful fallbackshub_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(themefurmanu)
register_furmanu_fonts()
```

# themefurmanu

The `themefurmanu` package provides a custom `ggplot2` theme and font registration functions using Furman University’s visual identity.

Note that this is not an official package from Furman University. It is a personal project to help with visual consistency in data visualizations across teams at Furman that use `R`.

## Installation

```r
# From GitHub
devtools::install_github("scottamoore/themefurmanu")
```

## The themefurmanu Approach: One Function, Many Possibilities

Instead of learning multiple function names like `scale_color_viridis_d()`, `scale_color_brewer()`, etc., themefurmanu uses **two main functions** with parameters:

- `scale_color_furmanu(palette = "name", discrete = TRUE/FALSE, ...)`
- `scale_fill_furmanu(palette = "name", discrete = TRUE/FALSE, ...)`

This approach is more intuitive and consistent across all use cases.

## Quick Start

```{r basic-example, fig.width=8, fig.height=5}
library(ggplot2)
library(themefurmanu)

# Basic scatter plot with categorical colors
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  scale_color_furmanu(palette = "main", discrete = TRUE) +
  theme_furmanu() +
  labs(
    title = "Fuel Efficiency by Car Weight",
    subtitle = "Colored by number of cylinders",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon",
    color = "Cylinders",
    caption = "Source: mtcars dataset"
  )
```

## Available Palettes

View all available palettes with their intended use cases:

```{r show-palettes, fig.width=8, fig.height=6}
show_furmanu_palettes()
```

```{r list-palettes}
# Get detailed information about palettes
list_palettes()
```

## Palette Types and Usage Guide

### Categorical Data (Discrete Groups)

For discrete categories like species, groups, or factors:

```{r categorical-example, fig.width=8, fig.height=5}
# Use the "main" palette for up to 6 categories
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_furmanu(palette = "main", discrete = TRUE) +
  theme_furmanu() +
  labs(
    title = "Iris Species Comparison",
    subtitle = "Using categorical color palette",
    x = "Sepal Length", y = "Sepal Width"
  )
```

### Sequential Data (Continuous, One Direction)

For continuous data that goes from low to high:

```{r sequential-example, fig.width=8, fig.height=5}
# Use sequential palettes for continuous data
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  scale_fill_furmanu(palette = "sequential1", discrete = FALSE) +
  theme_furmanu() +
  labs(
    title = "Old Faithful Eruption Patterns",
    subtitle = "Using sequential color palette",
    x = "Waiting Time (min)", y = "Eruption Duration (min)",
    fill = "Density"
  )
```

### Divergent Data (Continuous, Two Directions)

For data that diverges from a meaningful center point:

```{r divergent-example, fig.width=8, fig.height=5}
# Create divergent data
mtcars$mpg_diff <- mtcars$mpg - mean(mtcars$mpg)

ggplot(mtcars, aes(x = wt, y = hp, color = mpg_diff)) +
  geom_point(size = 4) +
  scale_color_furmanu(
    palette = "divergent1", 
    discrete = FALSE, 
    midpoint = 0  # This is key for divergent scales!
  ) +
  theme_furmanu() +
  labs(
    title = "Car Performance vs Fuel Efficiency",
    subtitle = "Color shows deviation from average MPG",
    x = "Weight (1000 lbs)", y = "Horsepower",
    color = "MPG\nDifference"
  )
```

## Theme Variants

### Standard Theme
```{r standard-theme, fig.width=7, fig.height=4}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(color = furmanu_cols("purple"), alpha = 0.7) +
  theme_furmanu() +
  labs(title = "Standard Furman Theme")
```

### Presentation Theme
```{r presentation-theme, fig.width=7, fig.height=4}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(color = furmanu_cols("purple"), alpha = 0.7) +
  theme_furmanu_presentation() +
  labs(title = "Presentation Theme")
```

### Minimal Theme
```{r minimal-theme, fig.width=7, fig.height=4}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(color = furmanu_cols("purple"), alpha = 0.7) +
  theme_furmanu_minimal() +
  labs(title = "Minimal Theme")
```

## Advanced Features

### Color Accessibility
```{r accessibility-check}
# Check contrast ratios
check_contrast("main")

# Find accessible combinations
head(accessible_combinations())
```

### Palette Export
```{r palette-export}
# Export as CSS variables
export_palette("main", format = "css")

# Export as JSON
export_palette("divergent1", format = "json")
```

## Tips for Effective Use

1. **Start with palette exploration**: Use `show_furmanu_palettes()` to see all options
2. **Choose the right type**: 
   - Categorical → "main" or "cool"
   - Sequential → "sequential1", "sequential2", "sequential3"  
   - Divergent → "divergent1", "divergent2", "divergent3"
3. **Use `discrete = FALSE` for continuous data**
4. **Add `midpoint = 0` for divergent continuous scales**
5. **Check accessibility**: Use `check_contrast()` for important visualizations
6. **Match theme to context**: Use `theme_furmanu_presentation()` for slides

## Getting Help

```r
# View comprehensive examples
open_furmanu_demo()

# Get help on any function
?scale_color_furmanu
?theme_furmanu
?show_furmanu_palettes

# List all available palettes
list_palettes()

# Get detailed palette information
palette_info("main")
```

See also `open_furmanu_demo()` to view the comprehensive example Quarto file with detailed tutorials and advanced techniques.
